Memory for television episodes preserves event content while introducing new across-event similarities

### 2020-02-03 paper/main.tex
working on intro

Although our experiences most often change gradually, they also occasionally change suddenly~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) during an experience help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within event) and transient (across event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated the hippocampus and the medial prefrontal cortex as playing a critical role in transforming experiences into stuctured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further examine whether 1) the \textit{precision} with which a participant recounts each event and 2) the \textit{distinctiveness} each recall event is (relative to the other recalled events) relates to their overall memory performance.  Last, we identify networks of  brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode, and how participants would later recount the episode.

\section*{Results}
To characterize the shape of the \textit{Sherlock} episode and participants' subsequent recountings of its unfolding, we used a topic model~\citep{BleiEtal03} to discover the latent themes in the episode's dynamic content.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening; whether the scene took place indoors or outdoors; the names of any characters on the screen; the names of any characters who were in focus in the camera shot; the names of characters who were speaking; the location where the scene took place; the camera angle (close up, medium, long, etc.); whether or not background music was present; and other similar details (for a full list of annotated features see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and,'' ``or,'' ``but,'' etc.) across all features and scenes as the ``vocabulary'' for the topic model.  We then concatenated the sets of words across all features contained in overlapping, 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision distributions over participants, for each video event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single video event.  Colored dots within the violin plots represent individual participants' recall precision for that event.  Video events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example video event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of video event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall precision distributions over participants, for each video event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single video event.  Colored dots within the violin plots represent individual participants' recall precision for that event.  Video events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example video event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of video event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:distinctiveness-detail}
\end{figure}

>>>>

Occasionally, this gradual ``drift" of our ongoing experience is punctuated by sudden changes, or ``shifts"~\citep[e.g., when we walk through a doorway; ][]{RadvZack17}.  Prior research suggests that these sharp transitions (termed \textit{event boundaries}) help to discretize our experiences (and their mental representations) into \textit{events}~\citep{RadvZack17, BrunEtal18, HeusEtal18b, ClewDava17, EzzyDava11, DuBrDava13}.  The interplay between the stable (within-event) and transient (across-event) temporal dynamics of an experience also provides a potential framework for transforming experiences into memories that distill those experiences down to their essence.  For example, prior work has shown that event boundaries can influence how we learn sequences of items~\citep{HeusEtal18b, DuBrDava13}, navigate~\citep{BrunEtal18}, and remember and understand narratives~\citep{ZwaaRadv98, EzzyDava11}. Prior research has implicated the hippocampus and the medial prefrontal cortex as playing a critical role in transforming experiences into stuctured and consolidated memories ~\citep{TompDava17}.

Here we sought to examine how the temporal dynamics of a ``naturalistic'' experience were later reflected in participants' memories.  We analyzed an open dataset that comprised behavioral and functional Magnetic Resonance Imaging (fMRI) data collected as participants viewed and then verbally recounted an episode of the BBC television series \textit{Sherlock}~\citep{ChenEtal17}.  We developed a computational framework for characterizing the temporal dynamics of the moment-by-moment content of the episode, and of participants' verbal recalls.  Specifically, we use topic modeling~\citep{BleiEtal03} to characterize the thematic conceptual (semantic) content present in each moment of the episode and recalls, and Hidden Markov Models~\citep{Rabi89, BaldEtal17} to discretize this evolving semantic content into events.  In this way, we cast naturalistic experiences (and recalls of those experiences) as geometric \textit{trajectories} that describe how the experiences evolve over time. Under this framework, successful remembering entails verbally ``traversing'' the content trajectory of the episode, thereby reproducing the shape (or essence) of the original experience.  Comparing the shapes of the topic trajectories of the episode and of participants' retellings of the episode then reveals which aspects of the episode were preserved (or lost) in the translation into memory.  We further introduce two novel metrics for assessing memory quality: the \textit{precision} with which a participant recounts each event and 2) the \textit{distinctiveness} of each recall event (relative to other recalled events).  We examine how these metrics relate to participants' overall memory performance, and discuss the ways in which they improve upon classic ``proportion-recalled" measures for analyzing naturalistic memory.  Last, we utilize our framework to identify networks of brain structures whose responses (as participants watched the episode) reflected the temporal dynamics of the episode, and how participants would later recount it.

\section*{Results}
To characterize the ``essence" of the \textit{Sherlock} episode and participants' subsequent recountings of its unfolding, we used a topic model~\citep{BleiEtal03} to discover the latent themes in the episode's dynamic content.  Topic models take as inputs a vocabulary of words to consider and a collection of text documents, and return two output matrices.  The first of these is a \textit{topics matrix} whose rows are topics (latent themes) and whose columns correspond to words in the vocabulary. The entries of the topics matrix define how each word in the vocabulary is weighted by each discovered topic.  For example, a detective-themed topic might weight heavily on words like ``crime,'' and ``search.''  The second output is a \textit{topic proportions matrix}, with one row per document and one column per topic.  The topic proportions matrix describes what mixture of discovered topics is reflected in each document.

\cite{ChenEtal17} collected hand-annotated information about each of 1000 (manually identified) scenes spanning the roughly 50 minute video used in their experiment.  This information included: a brief narrative description of what was happening; the location where the scene took place; the names of any characters on the screen; and other similar details (for a full list of annotated features, see \textit{Methods}).  We took from these annotations the union of all unique words (excluding stop words, such as ``and," ``or," ``but," etc.) across all features and scenes as the ``vocabulary" for the topic model.  We then concatenated the sets of words across all features contained in overlapping, 50-scene sliding windows, and treated each 50-scene sequence as a single ``document'' for the purpose of fitting the topic model.  Next, we fit a topic model with (up to) $K = 100$ topics to this collection of documents.  We found that 32 unique topics (with non-zero weights) were sufficient to describe the time-varying content of the video (see \textit{Methods}; Figs.~\ref{fig:schematic}, \topics).  Note that our approach is similar in some respects to Dynamic Topic Models~\citep{BleiLaff06} in that we sought to characterize how the thematic content of the episode evolved over time.  However, whereas Dynamic Topic Models are designed to characterize how the properties of \textit{collections} of documents change over time, our sliding window approach allows us to examine the topic dynamics within a single document (or video).  Specifically, our approach yielded (via the topic proportions matrix) a single \textit{topic vector} for each timepoint of the episode (we set timepoints to match the acquisition times of the 1976 fMRI volumes collected as participants viewed the episode).

@@

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/precision_detail}
  \caption{\small \textbf{Precision metric reflects completeness of recall.} \textbf{A.} Recall precision distributions over participants, for each video event.  Grey violin plots display kernel density estimates for the distribution of recall precision scores for a single video event.  Colored dots within the violin plots represent individual participants' recall precision for that event.  Video events are ordered along the $x$-axis by the average precision with which they were remembered.  \textbf{B.} The set of ``Narrative Details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example video event (22) identified by the HMM.  Each action or feature is highlighted in a different color.  \textbf{C.} A subset of the sentences comprising the most precise (P17) and least precise (P6) participants' recalls of video event 22.  Descriptions of specific actions or features reflecting those highlighted in panel B are highlighted in the corresponding color.}
  \label{fig:precision-detail}
\end{figure}

\begin{figure}[tp]
  \centering
  \includegraphics[width=1\textwidth]{figs/distinctiveness_detail}
  \caption{\small \textbf{Distinctiveness metric reflects specificity of recall.} \textbf{A.} Recall distinctiveness distributions over participants, for each video event.  Kernel density estimates for each video event's distribution of recall distinctiveness scores, analogous to Fig.~\ref{fig:precision-detail}A.  \textbf{B.} The set of ``Narrative Details" video annotations (generated by \citealp{ChenEtal17}) for scenes comprising an example video event (19) identified by the HMM.  \textbf{C.} The sentences comprising the most distinctive (P13) and least distinctive (P11) participants' recalls of video event 19.  Descriptions of recalled content specific to video event 19 (as opposed to that combining surrounding events) are highlighted.}
  \label{fig:distinctiveness-detail}
\end{figure}
